Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)

### 2024-12-02 paper/paper.qmd
Update paper.qmd

---
title: "Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)"
subtitle: "Insights from 2002 to 2023"
author: Dong Jun Yoon
thanks: "Code and data are available at: https://github.com/DJY1231/Exploring_Income_Inequality_in_Canada_During_Major_Recessions.git"
date: today
date-format: long
abstract: "This paper examines income inequality in Canada from 2002 to 2023 using data from the World Inequality Database (WID) and Statistics Canada. The analysis explores trends in income inequality and inflation, assessing whether these disparities have increased or decreased over time. It also compares the impacts of the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023) on income disparities. Through data analysis and regression modeling, this study investigates these trends in depth. Understanding these patterns is crucial for policymakers and the government to develop effective strategies aimed at reducing inequalities within the country. This report ultimately seeks to contribute to efforts promoting social and economic equity in Canada."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction
Income inequality is a significant issue both in Canada and all around the world. Income inequality refers to the difference in how income is distributed among the population, highlighting significant disparities in financial well-being. Unlike wealth inequality, which considers the value of assets such as homes, stocks, and other holdings, income inequality specifically examines earnings from sources like wages, salaries, savings interest, stock dividends, rental income, and profits from the sale of assets. This distinction is essential for understanding the multifaceted nature of economic disparities. Gaining a deeper understanding of income inequality can equip policymakers and researchers with valuable insights to develop strategies aimed at addressing this pervasive problem.

@@

@@

# Plot with smoothed lines, moving averages, and facets
ggplot(filtered_data, aes(x = Year, y = `Income Share`, color = Social_Class, group = Social_Class)) +
  geom_point(size = 2) + # Add points for individual data
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", alpha = 0.8) + # Smoothed lines for trends
  geom_line(aes(y = Moving_Avg), size = 1.2, linetype = "solid") + # Moving average line
  facet_wrap(~Country + Social_Class, scales = 'free_y') + # Facet by Country and Social Class
  scale_x_continuous(breaks = unique(filtered_data$Year), labels = as.character(unique(filtered_data$Year))) +
  labs(title = "Income Share during Covid-19 (2020-2023)",
       x = "Year",
       y = "Income Share (%)",
       color = "Social Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", # Adjust legend position
        strip.text = element_text(size = 10, face = "bold")) # Style facet labels

```
Figure 3: Income Share by Social Class during the Covid-19: 2020 to 2023 in Canada.

@@

## Model set-up
The multiple linear regression model is as follows: 
$$
\begin{align*}
\text{Share_of_total_income}(Y) &= \beta_0 + \beta_1 \cdot \text{Year} + \beta_2 \cdot \text{Social_Class} \\
&\quad + \beta_3 \cdot \text{Period} + \beta_4 \cdot (\text{Year} \times \text{Social_Class}) \\
&\quad + \beta_5 \cdot (\text{Social_Class} \times \text{Period}) + \varepsilon
\end{align*}
$$
### Model justification
We aim to model the relationship between the response variable Y, which represents the share of income for each social class, using the following components:

@@

# Display the regression results in a table with all combinations included
stargazer(model, type = "text", title = "Regression Results: Income Inequality During Great Recession and COVID-19", 
          dep.var.labels = "Share of Total Income",
          covariate.labels = c("Year", 
                               "Top 1% Income", 
                               "Top 10% Income", 
                               "Bottom 50% Income",
                               "Great Recession", 
                               "COVID-19",
                               "Other Period",
                               "Year x Top 1% Income", 
                               "Year x Top 10% Income", 
                               "Year x Bottom 50% Income",
                               "Top 1% Income x Great Recession", 
                               "Top 10% Income x Great Recession", 
                               "Bottom 50% Income x Great Recession",
                               "Top 1% Income x COVID-19", 
                               "Top 10% Income x COVID-19", 
                               "Bottom 50% Income x COVID-19",
                               "Top 1% Income x Other Period", 
                               "Top 10% Income x Other Period", 
                               "Bottom 50% Income x Other Period"),
          omit.stat = c("f", "ser"))

@@

## Posterior predictive check (PPC)

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

>>>>

---
title: "Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)"
subtitle: "Insights from 2002 to 2023"
author: Dong Jun Yoon
thanks: "Code and data are available at: https://github.com/DJY1231/Exploring_Income_Inequality_in_Canada_During_Major_Recessions.git"
date: today
date-format: long
abstract: "This paper examines income inequality in Canada from 2002 to 2023 using data from the World Inequality Database (WID) and Statistics Canada. The analysis explores trends in income inequality and inflation, assessing whether these disparities have increased or decreased over time. It also compares the impacts of the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023) on income disparities. Through data analysis and regression modeling, this study investigates these trends in depth. Understanding these patterns is crucial for policymakers and the government to develop effective strategies aimed at reducing inequalities within the country. This report ultimately seeks to contribute to efforts promoting social and economic equity in Canada."
format: pdf
number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction
Income inequality is a significant issue both in Canada and all around the world. Income inequality refers to the difference in how income is distributed among the population, highlighting significant disparities in financial well-being. Unlike wealth inequality, which considers the value of assets such as homes, stocks, and other holdings, income inequality specifically examines earnings from sources like wages, salaries, savings interest, stock dividends, rental income, and profits from the sale of assets. This distinction is essential for understanding the multifaceted nature of economic disparities. Gaining a deeper understanding of income inequality can equip policymakers and researchers with valuable insights to develop strategies aimed at addressing this pervasive problem.

@@

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(zoo) # For moving averages

@@

# Plot with smoothed lines, moving averages, and facets
ggplot(filtered_data, aes(x = Year, y = `Income Share`, color = Social_Class, group = Social_Class)) +
  geom_point(size = 2) + # Add points for individual data
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", alpha = 0.8, span = 0.75) + # Smoothed lines with adjusted span
  geom_line(aes(y = Moving_Avg), size = 1.2, linetype = "solid") + # Moving average line
  facet_wrap(~Country + Social_Class, scales = 'free_y') + # Facet by Country and Social Class
  scale_x_continuous(breaks = unique(filtered_data$Year), labels = as.character(unique(filtered_data$Year))) +
  labs(title = "Income Share during Covid-19 (2020-2023)",
       x = "Year",
       y = "Income Share (%)",
       color = "Social Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", # Adjust legend position
        strip.text = element_text(size = 10, face = "bold")) # Style facet labels
```
Figure 3: Income Share by Social Class during the Covid-19: 2020 to 2023 in Canada.

@@

## Model set-up
The multiple linear regression model is as follows: 
$$
\begin{aligned}
\text{Share\_of\_total\_income} (Y) &= \beta_0 + \beta_1 \cdot \text{Year} + \beta_2 \cdot \text{Social\_Class} \\
&\quad + \beta_3 \cdot \text{Period} + \beta_4 \cdot (\text{Year} \times \text{Social\_Class}) \\
&\quad + \beta_5 \cdot (\text{Social\_Class} \times \text{Period}) + \varepsilon
\end{aligned}
$$

### Model justification
We aim to model the relationship between the response variable Y, which represents the share of income for each social class, using the following components:

@@

# Generate the stargazer table and filter out the citation
# Capture the stargazer output in a variable
stargazer_output <- capture.output(
  stargazer(
    model, type = "text", title = "Regression Results: Income Inequality During Great Recession and COVID-19", 
    dep.var.labels = "Share of Total Income",
    covariate.labels = c("Year", 
                         "Top 1% Income", 
                         "Top 10% Income", 
                         "Bottom 50% Income",
                         "Great Recession", 
                         "COVID-19",
                         "Other Period",
                         "Year x Top 1% Income", 
                         "Year x Top 10% Income", 
                         "Year x Bottom 50% Income",
                         "Top 1% Income x Great Recession", 
                         "Top 10% Income x Great Recession", 
                         "Bottom 50% Income x Great Recession",
                         "Top 1% Income x COVID-19", 
                         "Top 10% Income x COVID-19", 
                         "Bottom 50% Income x COVID-19",
                         "Top 1% Income x Other Period", 
                         "Top 10% Income x Other Period", 
                         "Bottom 50% Income x Other Period"),
    omit.stat = c("f", "ser")
  )
)

# Remove lines containing "Please cite as" or citation messages
filtered_output <- stargazer_output[!grepl("Please cite as:", stargazer_output)]

# Print the filtered output
cat(filtered_output, sep = "\n")

@@

## Posterior predictive check (PPC)
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
